---
title: "Module 5 Project Report"
author: "stop_drop_and_code"
date: "2022-10-10"
output: html_document
bibliography: Mod5project.bib
---
```{r, include=FALSE}
library(tidyverse)
library(knitr)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
It has been previously established that body temperature and thermoregulation play integral roles in maintaining homeostasis of the body during forms of physical activity. @Lim2008human reports that metabolic rate increases drastically during physical activity. Furthermore, this increase in metabolic rate causes an increase in heat production. However, thermoregulation allows us to continue to function and sustains physiological processes. Hyperthermia may also be utilized by the central nervous system to act as a limiting mechanism to prevent damage to the body. This experiment investigates the effect of physical activity on body temperature in humans using temperature sensors. 

# Methods
The effect of physical activity on body temperature was assessed by collecting readings of axillary body temperature readings during a resting state and after a period of physical activity. The resting state was defined as having a resting heart rate while sitting on a couch. The active state was defined as being five minute of jumping jacks. Temperature in Celsius was recorded for 100 seconds during the resting state and after the active state in the axillary (armpit) region of each subject. An Arduino unit was used as a prototype along with a DHT11 temperature sensor to measure temperature data. 

# Results
```{r}
library(tidyverse)
dat.f <- list.files(pattern = "couch|active.csv") #find the files

dat.l <- list() #make an empty list

for(i in dat.f){
  met.dat<- unlist(strsplit(i,"_")) #split of the filename string
  who <- met.dat[1] #store the first element
  activity <- gsub(".csv","",met.dat[2]) #store second element, dropping CSV with gsub()
  dat.l[[i]]<- read_csv(i)%>%
    mutate(who=who,activity=activity)#read, then store data with additional columns of metada in list as position unique to i
}
dat <- do.call(rbind,dat.l) #combine the list into table
dat%>%
  ggplot()+geom_boxplot(aes(x=activity,y=degC))+facet_grid(.~who) + labs(caption="Figure 1. Mean Axial Body Temperature Between Resting and Active States by Subject") + theme(plot.caption=element_text(hjust=0.5) )
```

## Results Summary

# Discussion

# Author Contributions

Rachel R: Arduino setup and code

Rachel Z:

Jordan:

Michael: Data collection and production of .csv files, Introduction, Results, Methods

# References
