---
title: "Module 7 Report"
author: "stop-drop-and-code"
date: "2022-11-27"
output: html_document
bibliography: BIOL3140.bib
---

### Introduction



### Methods

*Do class-wide force-angle data for isometric MVC accord to a typical FL relationship both under control and fatigued conditions?*

To gather force-angle data, a DAQ consisting of a 40-kg load cell, HX711 amplifier, and an Arduino Uno microcontroller were used. The system was not calibrated because this report will be analyzing normalized force. Each group member pulled on a dowel connected to the load cell at 12 different angles for 30 seconds (45-157.5 degrees at 11.25 degree intervals), flexing at the elbow to produce MVCs. A ganiometer was attached to the upper arm to measure angles. Twelve angles were recorded without any additional exercise as a control, and 12 were recorded after isokinetic eccentric exercise dropping a backpack for 3 minutes to serve as a fatiguing condition. After waiting 2 minutes following the exercise, the fatigue experiments were conducted in the same way as the control.

These data were then analyzed to visualize force-length relationships in both conditions. First, data were normalized by taking the maximum force in each recording across all angles for each experiment for each individual, and dividing this Fmax by the Fmax for the experiment.

*Is there a significant shift between θmax between the control and fatigue FL relationships?*

### Results
```{r,include=FALSE}
library(tidyverse)
library(MuMIn)
library(geiger)
library(knitr)
```
```{r,include=FALSE}
#compiling metadata
f<- list.files("Project 8 data",full.names = T)
f.l <- list()
angs<- seq(45,168.75,length.out = 12)
for(i in f){
  f.i<-read_delim(i,delim = " ",show_col_types = FALSE, col_names = c("Reading","Force","Unit"))
  m <- unlist(strsplit(i,"_"))
  subject<-m[2]
  angle<-(angs[which.min(abs(angs-as.numeric(m[3])))])
  experiment<-gsub("\\..+","",m[4])
  f.l[[i]]<-f.i%>%
    mutate(subject=subject,angle=angle,experiment=experiment,)
}
dat<-do.call(rbind,f.l)
```
```{r,include=FALSE}
#finding max of each recording
dat_recording_max<-dat%>%
  group_by(subject,angle,experiment)%>%
  summarize(recording_max=max(abs(Force)))%>%
  print()
#finding max of each experiment
dat_Fmax<-dat_recording_max%>%
  group_by(subject,experiment)%>%
  summarize(Fmax=max(recording_max))%>%
  print()
#normalizing the max values
dat_Fmax_norm<-dat_recording_max%>%
    left_join(dat_Fmax)
dat_Fmax_norm<-dat_Fmax_norm%>%
  group_by(angle,subject,experiment)%>%
  summarize(Fmax_norm=(recording_max/Fmax))%>%
  print()
```
```{r,fig.cap="Figure 1. Normalized Fmax Values vs. Angle", echo=FALSE}
#plot of normalized maxes by experiment
dat_Fmax_norm%>%
  ggplot(aes(x=angle,y=Fmax_norm,col=experiment))+geom_point()
```
```{r,include=FALSE}
#Do class-wide force-angle data for isometric MVC accord to a typical FL relationship both under control and fatigued conditions?
dat_Fmax_norm_mean<-dat_Fmax_norm%>%
  group_by(angle,experiment)%>%
  summarize(Fmax_norm_mean=mean(Fmax_norm))
```
```{r,fig.cap="Figure 2. Mean Normalized Fmax vs. Angle", echo=FALSE}
#plot of means of normalized Fmax for each angle and experiment
dat_Fmax_norm_mean%>%
  ggplot(aes(x=angle,y=Fmax_norm_mean,col=experiment))+geom_point()+labs(y="Mean Normalized Fmax", x="Angle")
```
```{r,include=FALSE}
#time to model
poly.m2 <- lm(Fmax_norm~poly(angle,2),data=dat_Fmax_norm) #second order
poly.m3 <- lm(Fmax_norm~poly(angle,3),data=dat_Fmax_norm) #third order
poly.m4 <- lm(Fmax_norm~poly(angle,4),data=dat_Fmax_norm) #fourth order
#which model fits best for each experimental condition for each subject
poly.aic<-AICc(poly.m2,poly.m3,poly.m4) 
print(poly.aic) #2nd order, poly.m2, has the best fit
AICs <- dat_Fmax_norm%>%
  group_by(subject,experiment)%>%
  summarize(
    m2=AICc(lm(Fmax_norm~poly(angle,2))), #second order
    m3=AICc(lm(Fmax_norm~poly(angle,3))), #third order
    m4=AICc(lm(Fmax_norm~poly(angle,4))) #fourth order
  )%>%
  pivot_longer(m2:m4,names_to="model",values_to="AICc")%>%
  print()
```
```{r,include=FALSE}
#predicting values based on best fitting model for 1000 angles
x.pred<-seq(45,157.5,length.out=1000)
fits <- dat_Fmax_norm%>%
  group_by(subject,experiment)%>%
  summarize(
    m2=predict(lm(Fmax_norm~poly(angle,2)),newdata=data.frame(angle=x.pred)), #second order
    m3=predict(lm(Fmax_norm~poly(angle,3)),newdata=data.frame(angle=x.pred)), #third order
    m4=predict(lm(Fmax_norm~poly(angle,4)),newdata=data.frame(angle=x.pred)) #fourth order
  )%>%
  pivot_longer(m2:m4,names_to="model")%>%
  group_by(subject,experiment,model)%>%
  summarize(theta_max=x.pred[which.max(value)])%>%
  print()
#combining AICs and fits
best.models<-fits%>%
  left_join(AICs)%>%
  group_by(subject,experiment)%>%
  mutate(best=AICc==min(AICc))%>%
  filter(best==TRUE)%>%
  select(-best)%>%
  print()
```
```{r,include=FALSE}
#Is there a significant difference between control and fatigue theta_max? 
anova_thetamax<-anova(lm(theta_max~experiment,best.models))
```
```{r, echo=FALSE}
kable(anova_thetamax,caption="Table 1. Anova of Theta Max")
```
```{r,include=FALSE}

#Finding mean and SEM of shift in theta_max
theta_max<-best.models%>%
  pivot_wider(id_cols=subject,names_from = experiment,values_from=theta_max)%>%
  mutate(shift=fatigue-control)%>%
  ungroup()%>%
  summarize(mean.shift=mean(shift,na.rm=TRUE),se.shift=sd(shift,na.rm=TRUE)/sqrt(length(shift)))
```
```{r, echo=FALSE}
kable(theta_max,caption="Table 2.Mean and SEM Theta Max ")
```

### Discussion

### Author Contributions

Rachel Ruggera (Methods)

Rachel Zhu (Results)

Michael Britt (Introduction, Bibliography)

Jordan Nunes