---
title: "Module 5 Project Report"
author: "stop_drop_and_code"
date: "2022-10-10"
output: html_document
bibliography: [Mod5project.bib]
---
```{r, include=FALSE}
library(tidyverse)
library(knitr)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
It has been previously established that body temperature and thermoregulation play integral roles in maintaining homeostasis of the body during forms of physical activity. @Lim2008human reports that metabolic rate increases drastically during physical activity. Furthermore, this increase in metabolic rate causes an increase in heat production. However, thermoregulation allows us to continue to function and sustains physiological processes. Hyperthermia may also be utilized by the central nervous system to act as a limiting mechanism to prevent damage to the body. This experiment investigates the effect of physical activity on body temperature in humans using temperature sensors. 

# Methods
The effect of physical activity on body temperature was assessed by collecting readings of axillary body temperature readings during a resting state and after a period of physical activity. The resting state was defined as having a resting heart rate while sitting on a couch. The active state was defined as being five minute of jumping jacks. Temperature in Celsius was recorded for 100 seconds during the resting state and after the active state in the axillary (armpit) region of each subject. An Arduino unit was used as a prototype along with a DHT11 temperature sensor to measure temperature data. 

# Results
```{r, include=FALSE}
library(tidyverse)
dat.f <- list.files(pattern = "couch|active.csv") #find the files

dat.l <- list() #make an empty list

for(i in dat.f){
  met.dat<- unlist(strsplit(i,"_")) #split of the filename string
  who <- met.dat[1] #store the first element
  activity <- gsub(".csv","",met.dat[2]) #store second element, dropping CSV with gsub()
  dat.l[[i]]<- read_csv(i)%>%
    mutate(who=who,activity=activity)#read, then store data with additional columns of metada in list as position unique to i
}
dat <- do.call(rbind,dat.l) #combine the list into table
```
```{r, echo=FALSE}
dat%>%
  ggplot()+geom_boxplot(aes(x=activity,y=degC))+facet_grid(.~who) + labs(caption="Figure 1. Mean Axial Body Temperature Between Resting and Active States by Subject") + theme(plot.caption=element_text(hjust=0.5) )
```

### Results Summary

The mean active body temperatures for Jordan, Michael, and RachelZ are all higher than their mean resting body temperatures. The mean active body temperature vs mean resting body temperature for RachelR does not follow this trend. RachelR's mean active body temperature appears to be slightly lower than her mean resting temperature. 

# Discussion

*Does body temperature vary in human subjects according to activity level?*

From our results we can conclude that body temperature does vary in human subjects according to activity level. @gleeson1998temperature reports that during strenuous exercise, the body may produce upwards of 1000 W of heat. Some of that heat is stored, ultimately raising the body's core temperature by a few degrees. There was a discrepancy in results with RachelR's body temperature, however we conclude that this discrepancy is a result of experimental error, possibly from incorrect placement of the temperature sensor, or from interference with the clothes RachelR was wearing when her temperature was taken. These findings are congruent to our results, and support the idea that body temperature rises as a result of physical activity. 

# Author Contributions

Rachel R: Arduino setup and code

Rachel Z: Results Summary and Discussion

Jordan:

Michael: Data collection and production of .csv files, Introduction, Results, Methods

# References
